Semantic Linearity in a Lazy Optimising Compiler

### 2025-02-19 icfp.tex
environment expansion keeps env outerwards
fix useful lemma
drop unnecessary \x from typing env in x=e|x
fix typo
bad lemma update
downarrow^1_ fix Lemma C3
Fix Case reduction auxiliary Sigma things
Tweaks and fixes to operational semantic
No, the e' in the Sigma is not substituted
There are two cases:

In the "e is in WHNF" we don't need that induction hypothesis to deduce
the result

In the "e is not in WHNF", the resources introduced in e' by the case
are "irrelevant", so we can move the \D bound variables to linear and
wipe that from the environment to reconstruct (e, e') and apply the
ind

\subsection{Language Syntax and Operational Semantics}

@@

The (small-step) operational semantics of Linear Core are given in
Figure~\ref{fig:linear-core-operational-semantics}. We use call-by-name
evaluation for Linear Core as it captures the non-strict semantics in which
our type system understands linearity, while being simpler to reason about than
call-by-need operational semantics.
% which is traditionally modelled with a
% mutable heap to store \emph{thunks} and the values they are overwritten with.
Furthermore, linear function applications, even in a \emph{call-by-need} system, can
usually be reduced \emph{call-by-name} as the function argument is guaranteed to
be used exactly once (thus avoiding unnecessarily allocating memory on the heap
for a redundant \emph{thunk}).
%
Specifically, function applications are reduced by the standard
\emph{call-by-name} $\beta$-reduction, substituting the argument by
occurrences of the binder in its body, case expressions evaluate their
scrutinee to WHNF, substituting the result by the case binder and constructor
arguments for pattern-bound bound variables matching on that same constructor.

\input{language-v4/OperationalSemantics}

@@

@@

\section{Instrumented Operational Semantics}

@@

\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Gamma;\Delta \vdash (\Theta \mid \Lambda p.e) \Downarrow (\Theta \mid
    \Lambda p.e) : \forall p . \sigma , \Sigma }
    \\[1.5em]
    \infer[]
    {\Gamma ; \Delta \vdash (\Theta \mid  e) \Downarrow (\Theta' \mid
    \Lambda p.e') : \forall p . \sigma , \Sigma \quad \Gamma ;\Delta \vdash
    (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' \mid v) : \sigma, \Sigma}
    {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
    \mid v) : \sigma , \Sigma  }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma ;\Delta \vdash (\Theta \mid \lambda x:_\pi \sigma . e)
    \Downarrow (\Theta \mid \lambda x:_\pi\sigma .e) : \sigma
    \rightarrow_\pi \tau,\Sigma }\\[1.5em]
    \infer[]
    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
    \lambda y:_\pi \sigma . e') : \sigma \rightarrow_\pi \tau , x:\sigma, \Sigma
    \quad
    \Gamma ; \Delta \vdash (\Theta' , y:_{\pi} \sigma = x \mid e') \Downarrow (\Theta''
    \mid v) : \tau , \Sigma  }
    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
    \mid v) : \tau , \Sigma }\\[1.5em]

       \infer[]
    {\Gamma , x:_{\Delta'} \sigma ; \Delta  \vdash (\Theta \mid e)
    \Downarrow (\Theta' \mid v) : \sigma , \Sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
    \Downarrow (\Theta' , x:_{\Delta'} \sigma = v \mid v) : \sigma, \Sigma}
    \\[1.5em]
           \infer[]
    {\Gamma ;\Delta  \vdash (\Theta \mid e)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_1 \sigma = e \mid x)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    \\[1.5em]

@@

    \infer[]
    { K : \ov{ \sigma_i
    \rightarrow_\omega \sigma_j \lolli} T
    \\
    \Gamma, \ov{x_i :_\omega \sigma_i} , z:_{\Delta'} T ;\Delta , \ov{x_j :_1 \sigma_j} \vdash (\Theta \mid e) \Downarrow (\Theta' :
    K~\ov{x_i}) : T , e' : \tau ,\Sigma \\
    \Gamma ;\Delta \vdash (\Theta' \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z]
    ) \Downarrow (\Theta'' \mid v) :\tau , \Sigma }
    {\Gamma ;\Delta \vdash (\Theta \mid
    \ccase{e}{z{:}_{\Delta'}T~\{{\dots,K~\ov{y_i} \to e', {\dots} }\}})
    \Downarrow (\Theta'' \mid  v) : \tau , \Sigma}\\[1.5em]

    \infer[]
    {K : \ov{ \sigma_i
    \rightarrow_\omega \sigma_j \lolli} T\\
    \Gamma, z:_{\Delta'} T ;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
    K~\ov{x_i}) : T , e' , \Sigma \quad
    \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~\ov{x_i}/z])
    \Downarrow (\Theta'' \mid v) : \tau,\Sigma }
    {\Gamma ; \Delta \vdash (\Theta \mid 
    \ccase{e}{z{:}_{\Delta'}T~\{\dots , \_\to
    e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
    \\[1.5em]
    % \infer[]
    % {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % % \Downarrow \Theta_n : v_n \quad
    % (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\omega \sigma_i =
    % e_i} \mid e' )\Downarrow (\Theta'
    % \mid v) : \tau
    % }
    % {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'})
    % \Downarrow (\Theta' \mid v) : \tau}

@@

 \begin{definition}[Environment Expansion]
   We rely on an expansion of a runtime environment $\Theta$ for the evaluation of a term $e$ into a (typed) term, written $[\Theta \mid e]$, defined inductively as follows:
   \[
     \begin{array}{lcl}
       \left[ \cdot \mid e \right] & \triangleq & e\\
       \left[ \Theta , x:_1\sigma = e' \mid e \right] & \triangleq  & (\lambda x_1:\sigma . [\Theta \mid e])~e'\\
       \left[\Theta , x:_\Delta \sigma = e' \mid e \right] & \triangleq & \llet{x:_\Delta \sigma = e'}{[\Theta \mid e]}
    \end{array}
   \]
 \end{definition}

@@

 \begin{lemma}
   Let $\G ; \D , \ov{x_i :_{\Delta_i} \tau_i} ; \ov{x_j :_1 : \tau_j} \vdash  e : \sigma$, we have that
   $\G ; \D , \ov{\Delta_i} \vdash [ \ov{x_i :_{\Delta_i} \tau_i = e_i} , \ov{x_j :_1 : \tau_j = e_j} \mid e] : \sigma$
   \end{lemma}

>>>>

\subsection{Language Syntax}

@@

% The (small-step) operational semantics of Linear Core are given in
% Figure~\ref{fig:linear-core-operational-semantics}. We use call-by-name
% evaluation for Linear Core as it captures the non-strict semantics in which
% our type system understands linearity, while being simpler to reason about than
% call-by-need operational semantics.
% % which is traditionally modelled with a
% % mutable heap to store \emph{thunks} and the values they are overwritten with.
% Furthermore, linear function applications, even in a \emph{call-by-need} system, can
% usually be reduced \emph{call-by-name} as the function argument is guaranteed to
% be used exactly once (thus avoiding unnecessarily allocating memory on the heap
% for a redundant \emph{thunk}).
% %
% Specifically, function applications are reduced by the standard
% \emph{call-by-name} $\beta$-reduction, substituting the argument by
% occurrences of the binder in its body, case expressions evaluate their
% scrutinee to WHNF, substituting the result by the case binder and constructor
% arguments for pattern-bound bound variables matching on that same constructor.

% \input{language-v4/OperationalSemantics}

@@

We begin by presenting the operational semantics of $\lambda^\pi_\D$,
consisting of two lazy big-step semantics in the
style of Launchbury~\cite{10.1145/158511.158618}.

\subsection{Operational Semantics}

The first semantics, presented in Figure~\ref{fig:opsem}, captures
lazy evaluation but ignores linearity information. The second
semantics is instrumented with type information and is thus
\emph{linearity aware}, resulting in a stuck state when a linear
variable is used more than once. We show that for well-typed terms,
the two semantics coincide.

Both semantics are
equipped with an environment which assigns values to variables and is
mutated in order to express lazy evaluation. Terms are explicit about
their sharing (in the sense of shared reductions in lazy evaluation),
and so we perform a translation step that makes all sharing explicit
through let-binders (see Appendix~\ref{app:opsem}).
We write $\llet{\ov{x_i{:}_{\Delta_i}\sigma = e_i}}{e'}$ for the iterated
let-binding of variables $\ov{x_i}$.

\begin{figure}

  {\small
    \[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Theta : \Lambda p.e \Downarrow \Theta : \Lambda p.e}
    \qquad
    \infer[]
    {\Theta : e\Downarrow \Theta' : \Lambda p.e' \quad \Theta' :
    e'[\pi/p]\Downarrow \Theta'' : v}
    {\Theta : e~\pi\Downarrow \Theta'' : v }\\[1.5em]
    \infer[]
    {\,}
    {\Theta : \lambda x:_\pi \sigma . e \Downarrow \Theta : \lambda x:_\pi \sigma . e }
    \qquad
    \infer[]
    {\Theta : e \Downarrow \Theta' : \lambda y:_\pi \sigma . e' \quad
    \Theta' : e'[x/y] \Downarrow \Theta'' : v }
    {\Gamma : e~x \Downarrow \Theta'' : v}\\[1.5em]
    \infer[]
    {\Theta : e \Downarrow \Theta' : v}
    {(\Theta , x:_\omega \sigma = e) : x \Downarrow (\Theta',x:_\omega
    \sigma = v) : v}
    \qquad
    \infer[]
    {(\Theta,x:_\omega \sigma = e) : e' \Downarrow \Theta' : v}
    {\Theta : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Theta' : v
    }\\[1.5em]
    \infer[]
    {\,}
    {\Theta : K~\ov{x_i} \Downarrow \Theta : K~\ov{x_i}}
    \qquad
    \infer[]
    {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
    \Theta' : e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
    {\Theta :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i} \to e', {\dots}}\}}
    \Downarrow \Theta'' : v}\\[1.5em]
    \infer[]
    {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
    \Theta' : e'[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
    {\Theta :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
    e'}\} \Downarrow \Theta'' : v}
    \qquad
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    (\Theta, \overline{x_i:_\omega \sigma_i = e_i}) : e' \Downarrow \Theta'
    : v
    }
    {\Theta : \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'}
    \Downarrow \Theta' : v}
    \end{array}
\]}
    \caption{Operational Semantics of $\lambda^\pi_\D$\label{fig:opsem}}
  \end{figure}

  The lazy big-step operational semantics is defined via the relation
$\Theta : e \Downarrow \Theta' : v$ where $e$ is an expression, $v$ a
value, $\Theta$ and $\Theta'$ are environments with bindings of the
form $x:_\omega\sigma = e$ which assigns the expression $e$ to the
variable $x$ of the given type.

We follow an approach
similar to that of \cite{cite:linearhaskell}, defining a
linearity-aware operational semantics in which linear variables are
erased from 
the runtime environment, ensuring that any term violating linearity
will result in a stuck state. The semantics is also type-aware (for
the purposes of proving type safety), defined by
the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing judgments
($\Gamma$ tracks unrestricted variables, $\Delta$ tracks linear
assumptions);
$\Theta$ and $\Theta'$ are runtime environments, consisting of bindings of the form
$x :_\pi \sigma = e$; $e$ is the expression being evaluated and $v$
its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
$e :_\pi \tau$ or $x :_\Delta \tau$ which are expressions in which
variables in $\Theta$ are used.

\begin{figure}
\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Gamma;\Delta \vdash (\Theta \mid \Lambda p.e) \Downarrow (\Theta \mid
    \Lambda p.e) : \forall p . \sigma , \Sigma }
    \\[1.5em]
    \infer[]
    {\Gamma ; \Delta \vdash (\Theta \mid  e) \Downarrow (\Theta' \mid
    \Lambda p.e') : \forall p . \sigma , \Sigma \quad \Gamma ;\Delta \vdash
      (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' \mid v) : \sigma[\pi/p], \Sigma}
    {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
      \mid v) : \sigma[\pi/p] , \Sigma  }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma ;\Delta \vdash (\Theta \mid \lambda x:_\pi \sigma . e)
    \Downarrow (\Theta \mid \lambda x:_\pi\sigma .e) : \sigma
    \rightarrow_\pi \tau,\Sigma }\\[1.5em]
    \infer[]
    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
    \lambda y:_\pi \sigma . e') : \sigma \rightarrow_\pi \tau , x:\sigma, \Sigma
    \quad
    \Gamma ; \Delta \vdash (\Theta' , y:_{\pi} \sigma = x \mid e') \Downarrow (\Theta''
    \mid v) : \tau , \Sigma  }
    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
    \mid v) : \tau , \Sigma }\\[1.5em]

       \infer[]
    {\Gamma ; \Delta  \vdash (\Theta \mid e)
    \Downarrow (\Theta' \mid v) : \sigma , \Sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
      \Downarrow (\Theta' , x:_{\Delta''} \sigma = v \mid v) : \sigma, \Sigma}
    \\[1.5em]
           \infer[]
    {\Gamma ;\Delta  \vdash (\Theta \mid e)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_1 \sigma = e \mid x)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    \\[1.5em]

    \infer[]
    { K : \ov{ \sigma_i
    \rightarrow_\omega \sigma_j \lolli} T
    \\
    \Gamma, \ov{y_i :_\omega \sigma_i} ;\Delta , \ov{y_j :_1 \sigma_j} \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
      K~\ov{x_i}) : T , e' : \tau ,\Sigma \\
    \Gamma ;\Delta \vdash (\Theta' \mid e'\ov{[x_i/y_{i,j}]}[K~\ov{x_i}/z]
    ) \Downarrow (\Theta'' \mid v) :\tau , \Sigma }
    {\Gamma ;\Delta \vdash (\Theta \mid
    \ccase{e}{z{:}_{\Delta'}T~\{{\dots,K~\ov{y_{i,j}} \to e', {\dots} }\}})
    \Downarrow (\Theta'' \mid  v) : \tau , \Sigma}\\[1.5em]

    \infer[]
    {K : \ov{ \sigma_i
    \rightarrow_\omega \sigma_j \lolli} T\\
    \Gamma ; \Delta, z{:}_1\s \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
    K~\ov{x_i}) : T , e', \Sigma \quad
    \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~\ov{x_i}/z])
    \Downarrow (\Theta'' \mid v) : \tau,\Sigma }
    {\Gamma ; \Delta \vdash (\Theta \mid 
    \ccase{e}{z{:}_{\Delta'}T~\{\dots , \_\to
    e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
    \\[1.5em]
    % \infer[]
    % {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % % \Downarrow \Theta_n : v_n \quad
    % (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\omega \sigma_i =
    % e_i} \mid e' )\Downarrow (\Theta'
    % \mid v) : \tau
    % }
    % {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'})
    % \Downarrow (\Theta' \mid v) : \tau}

   \end{array}
 \]
 \caption{Linearity-aware Operational Semantics~\label{fig:linopsem}}

\end{figure}

\subsection{Results}

@@

\section{Instrumented Operational Semantics}\label{app:opsem}

@@

\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Gamma;\Delta \vdash (\Theta \mid \Lambda p.e) \Downarrow (\Theta \mid
    \Lambda p.e) : \forall p . \sigma , \Sigma }
    \\[1.5em]
    \infer[]
    {\Gamma ; \Delta \vdash (\Theta \mid  e) \Downarrow (\Theta' \mid
    \Lambda p.e') : \forall p . \sigma , \Sigma \quad \Gamma ;\Delta \vdash
      (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' \mid v) : \sigma[\pi/p], \Sigma}
    {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
      \mid v) : \sigma[\pi/p] , \Sigma  }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma ;\Delta \vdash (\Theta \mid \lambda x:_\pi \sigma . e)
    \Downarrow (\Theta \mid \lambda x:_\pi\sigma .e) : \sigma
    \rightarrow_\pi \tau,\Sigma }\\[1.5em]
    \infer[]
    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
    \lambda y:_\pi \sigma . e') : \sigma \rightarrow_\pi \tau , x:\sigma, \Sigma
    \quad
    \Gamma ; \Delta \vdash (\Theta' , y:_{\pi} \sigma = x \mid e') \Downarrow (\Theta''
    \mid v) : \tau , \Sigma  }
    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
    \mid v) : \tau , \Sigma }\\[1.5em]

       \infer[]
    {\Gamma ; \Delta  \vdash (\Theta \mid e)
    \Downarrow (\Theta' \mid v) : \sigma , \Sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
      \Downarrow (\Theta' , x:_{\Delta''} \sigma = v \mid v) : \sigma, \Sigma}
    \\[1.5em]
           \infer[]
    {\Gamma ;\Delta  \vdash (\Theta \mid e)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_1 \sigma = e \mid x)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    \\[1.5em]

@@

    \infer[]
    { K : \ov{ \sigma_i
    \rightarrow_\omega \sigma_j \lolli} T
    \\
    \Gamma, \ov{y_i :_\omega \sigma_i} ;\Delta , \ov{y_j :_1 \sigma_j} \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
      K~\ov{x_i}) : T , e' : \tau ,\Sigma \\
    \Gamma ;\Delta \vdash (\Theta' \mid e'\ov{[x_i/y_{i,j}]}[K~\ov{x_i}/z]
    ) \Downarrow (\Theta'' \mid v) :\tau , \Sigma }
    {\Gamma ;\Delta \vdash (\Theta \mid
    \ccase{e}{z{:}_{\Delta'}T~\{{\dots,K~\ov{y_{i,j}} \to e', {\dots} }\}})
    \Downarrow (\Theta'' \mid  v) : \tau , \Sigma}\\[1.5em]

    \infer[]
    {K : \ov{ \sigma_i
    \rightarrow_\omega \sigma_j \lolli} T\\
    \Gamma ; \Delta, z{:}_1\s \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
    K~\ov{x_i}) : T , e', \Sigma \quad
    \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~\ov{x_i}/z])
    \Downarrow (\Theta'' \mid v) : \tau,\Sigma }
    {\Gamma ; \Delta \vdash (\Theta \mid 
    \ccase{e}{z{:}_{\Delta'}T~\{\dots , \_\to
    e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
    \\[1.5em]
    % \infer[]
    % {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % % \Downarrow \Theta_n : v_n \quad
    % (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\omega \sigma_i =
    % e_i} \mid e' )\Downarrow (\Theta'
    % \mid v) : \tau
    % }
    % {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'})
    % \Downarrow (\Theta' \mid v) : \tau}

@@

 \begin{definition}[Environment Expansion]
   We rely on an expansion of a runtime environment $\Theta$ for the evaluation of a term $e$ into a (typed) term, written $[\Theta \mid e]$, defined inductively as follows:
   \[
     \begin{array}{lcl}
       \left[ \cdot \mid e \right] & \triangleq & e\\
       \left[ \Theta , x:_1\sigma = e' \mid e \right] & \triangleq  & [\Theta \mid (\lambda x_1:\sigma . e)~e']\\
       \left[\Theta , x:_\Delta \sigma = e' \mid e \right] & \triangleq & [\Theta \mid \llet{x:_\Delta \sigma = e'}{e}]
    \end{array}
   \]
 \end{definition}

@@

 \begin{lemma}
   Iff $\G ; \D \vdash [ \Theta \mid e] : \sigma$
     then $\G , \Theta\uparrow^{\omega} ; \D , \Theta\uparrow^1 \vdash  (e, \Theta\downarrow^1_e) : \sigma \otimes \bigotimes(\Theta\downarrow^1_\sigma)$
   \end{lemma}
